Dynamical Stochastic Approximation Monte Carlo

### 2018-02-20 papers/histogram/sad.tex

\author{Jordan K. Pommerenck} \author{Tanner T. Simpson}
\author{Michael A. Perlin} \author{David Roundy}
\affiliation{Department of Physics, Oregon State University,
  Corvallis, OR 97331}

@@

In our work, we have developed a novel algorithm based on SAMC that
does not require user-defined inputs and therefore should be easily
applicable to a given system.  We call this method SAD (Dynamical
Stochastic Approximation), and will discuss it in detail in the methods
section. We compare it along with several histogram methods including TMMC,
WL, WL-TMMC, and SAMC.

@@

Here we employ a variety of broad-histogram methods.  We outline the
general workings of algorithm we developed in detail and summarize
algorithms that can be found elsewhere.  The following algorithms are
introduced and applied to the square-well fluid:  Transition
Matrix Monte-Carlo (TMMC), Wang-Landau (WL), Wang-Landau Transition
Matrix Monte-Carlo (WL-TMMC), Stochastic Approximation Monte-Carlo
Monte-Carlo (SAMC), and Dynamical Stochastic-Approximation (SAD).

@@

\subsection{SAD Algorithm}
The SAD (Dynamic Stochastic Approximation) is a version of the SAMC
Algorithm that attempts to dynamically choose the modification factor
rather than relying on system dependent parameters such as $t_0$ or
$\gamma_0$.  There is an immediate advantage of such an algorithm where
parameters are chosen independent of system size or type. SAMC can be
applied as a dynamic importance sampling algorithm that uses an update
factor to explore energy space~\cite{liang2007stochastic,
werlich2015stochastic, schneider2017convergence}.  The update factor is
defined in the original implementation~\cite{liang2007stochastic} in
terms of an arbitrary tunable parameter $t_0$.
\begin{align}
\gamma_{t}^{\text{SA}} = \frac{t_0}{\max(t_0,t)}
\end{align}
We argue that it is possible to set this parameter to a system
independent parameter such that $t_0 = N_E^2/t$ where $N_E$ is the
number of energies that have been found.  One of the energy found terms
can be thought of as a change in the entropy $\Delta S$.  The change in
the entropy can be thought of as a slope of the triangle made by
connecting the minimum important energy and the maximum entropy state
divided by the minimum temperature:
\begin{align}
\Delta S = \frac{E_{\text{max}}-E_{\text{min}}}{T_{\text{min}}}
\end{align}
This gives an update factor as follows:
\begin{align}
\gamma_{t}^{\text{SA}} = \frac{N_E\Delta S}{t} =
\frac{N_E}{t}\frac{E_{\text{max}}-E_{\text{min}}}{T_{\text{min}}}
\end{align}
The update factor has the same $1/t$ behavior as the original SAMC
algorithm; however now every time a new energy is discovered during the
course of the simulation, the update factor experiences a jump.
~\jpsays{Add in additional material for SAD!}

@@

For the square-well fluid, our simulations treat systems with a
well-width of $\text{ww} = 1.30$ and a filling fraction of either
$\text{ff} = 0.10, 0.20$ or $\text{ww} = 0.20$.  The simulation explores the
energy space of the system down to a reduced temperature of
$T_{\text{min}} = 0.2$.  All simulations lead to the minimum important
energy $E_{\text{min}}$ and maximum entropy state $S_{\text{max}}$
being calculated (with the exception of the WL-TMMC method where both
of these parameters are needed a priori).  In the sections below, we
examine systems with different particles numbers and filling fractions.

~\jpsays{Add N50 ff0.1 and N50 ff0.3 and remove sticky-wall!}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/lv/ww1_30-ff0_22-30x4-max-entropy-error.pdf}
  \caption{The average entropy error for each method is shown with
  SAD serving as the `golden'
  standard.}\label{fig:avgS-30x4}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/lv/ww1_30-ff0_22-60x8-max-entropy-error.pdf}
  \caption{The average entropy error for each method is shown with SAD serving as
  the `golden' standard.}\label{fig:avgS-60x8}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/lv/ww1_30-ff0_22-100x10-max-entropy-error}
  \caption{The average entropy error for each method is shown with
    WL-TMMC using a cutoff of 10$^{-10}$ serving as the
    standard.}\label{fig:avgS-100x10}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_20-N50-max-entropy-error.pdf}
  \caption{The average entropy error is shown with TMI acting as the
  `golden' standard.}\label{fig:N50-mean-error}
\end{figure}
\begin{figure}
\includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N500-max-entropy-error.pdf}
  \caption{The average entropy error is shown with TMI acting as the
  `golden' standard.}
\end{figure}

>>>>

\author{Jordan K. Pommerenck} \author{Michael A. Perlin}
\author{Tanner T. Simpson} \author{David Roundy}
\affiliation{Department of Physics, Oregon State University,
  Corvallis, OR 97331}

@@

In our work, we have developed several novel algorithms that do not
require user-defined inputs and therefore should be easily applicable
to a given system.  These methods, each of which will be discussed in
detail in the methods section, can also be made to satisfy theoretical
requirements, guaranteeing convergence. These algorithms are Transition
Matrix Initialization (TMI), Dynamic Stochastic Approximation (SAD),
and Stochastic Approximation Transition Matrix Monte Carlo (SA-TMMC).
TMI works by using logical conditionals about changes in the logrithm
of the density of states to control the weights of the simulation.  SAD
operates effectively the same as standard SA; however, we claim that
the user defined parameter can be set independent of the system type or
size.  SA-TMMC works by performing SA for a system-independent amount
of time in order to explore the entire energy space.  The simulation
then switches over to TMMC for the remainder of the simulation.  All of
these methods have no parameter inputs that depend on the type of
system or size of system under study.

@@

Here we employ a variety of broad-histogram methods.  We outline the
general workings of algorithm we developed in detail and summarize
algorithms that can be found elsewhere.  The following algorithms are
introduced and applied to the square-well fluid: Transition Matrix
Initialization (TMI), Transition Optimized Ensemble (TOE), Transition
Matrix Monte-Carlo (TMMC), Wang-Landau (WL), Wang-Landau Transition
Matrix Monte-Carlo (WL-TMMC), Dynamic Stochastic Approximation
Monte-Carlo (SAD), and Stochastic-Approximation Transition Matrix
Monte-Carlo (SA-TMMC).

\subsection{TMI2 Algorithm}

The second version of transition matrix initialization algorithm works
by looking at the logarithm of the density of states.  We begin at
high energies ($\epsilon_{S_{\max}}$ the energy that maximizes the
entropy) and work our way
downward.

\paragraph{Reaching $T_{\min}$:}
If the change in the logarithm of the density of states is greater
than the reciprocal of the minimum temperature, i.e. if
\begin{align}
  \ln\frac{D(\epsilon_i)}{D(\epsilon_{i-1})} &>
  \frac{\epsilon_{i-1} - \epsilon_i}{k_BT_{\min}}
  \\
  \frac{D(\epsilon_i)}{D(\epsilon_{i-1})} &>
  e^{\beta_{\max}(\epsilon_{i-1} - \epsilon_i)}
\end{align}
then the algorithm has reached the minimum temperature of interest.
In this case, we set the log weights at all lower energies using a
Boltzmann ratio relative to this energy.
\paragraph{Confident at $\epsilon_i$:} We are confident in our density
of states at energy $\epsilon_i$ provided two conditions are true.
\begin{enumerate}
\item The density of states must be decreasing as the energy decreases
  from its next-higher energy.
\item The change in density of states (relative to the next higher
  energy) must be smaller than a pessimistic estimate of the counting
  uncertainty at the energy we are examining:
  \begin{align}
    \frac{D(\epsilon_i)}{D(\epsilon_{i-1})} < \frac1{\sqrt{N_i}}
  \end{align}
  where $N_i$ is the number of ``pessimistic
  samples'' at $\epsilon_i$, the lower of the two energies under
  consideration.
\end{enumerate}
If both of these are true, we believe we know the density of states at
this energy well, and the weights are set by
\begin{align}
  \ln w(\epsilon_i) = -\ln{D}(\epsilon_i)
\end{align}
\paragraph{Reached $T_\text{converged}$:}
If this is not the case, $\epsilon_i$ is the lowest energy that is
adequately converged.  We set the weights for energies lower than this
to a Boltzmann ratio with a temperature determined using the slope of
a secant line on the $\ln w$ versus $\epsilon$ graph.
\begin{align}
  \beta_\text{secant} = \bigg(\frac{\ln{D}(\epsilon_{S_{\max}}) - \ln{D}
  (\epsilon_\text{pivot})}{\epsilon_{S_{\max}}-\epsilon_\text{pivot}}\bigg)
\end{align}
This ``slope'' is a positive quantity.  The weights at energies below
$\epsilon_\text{pivot}$ are given by
\begin{align}
  \ln w(\epsilon) = \ln w(\epsilon_\text{pivot}) -
  \beta_\text{secant}(\epsilon - \epsilon_\text{pivot})
\end{align}
indicating that there are higher weights at lower energies, since
$\beta$ is positive.  We have one additional caveat: we examine the
lower energies for which we have samples (``pessimistic'' samples),
and if we find any energy at which the density of states is greater
than the inverse of the weight, we adjust $\beta$ to cause it to pass
through that point.  This ensures that the product of the weight with
the estimated density of states is always less than or equal one for
energies below $\epsilon_\text{converged}$.

\subsection{TMI3 Algorithm}
The TMI$_3$ algorithm is like the TMI$_2$ algorithm, with one
difference, which is that instead of a secant line being used to
estimate the converged temperature, we use a tangent line, estimated
using the lowest two ``converged'' energies.

\subsection{TOE3 Algorithm}
The TOE$_3$ method desribed by Trebst et.
al.~\cite{trebst2004optimizing} uses the metric of round trips defined
by a walker visiting the extreme energies for the density of states.
The rate for the walker is dependent of the diffusivity $D(\epsilon)$
at a given energy $E$.  The goal of the method is to maximize the
number of round trips that the Monte-Carlo simulation makes.  The
optimized density of states is given by the following:
\begin{align}
  n_{w}^{(opt)} = \frac{1}{\sqrt{D(E)}}
\end{align}
where unlike in the original paper detailing the method, we compute the
diffusivity directly from the transition matrix normalized with
canonical weights.  The optimal ensemble therefore relates the density
of states to the inverse of the square root of the diffusivity.
\begin{align}
  D(E) = \mid (T(\Delta\epsilon)^2 - (T(\Delta\epsilon))^2) \mid
\end{align}

@@

\subsection{SAD Algorithm}
The SAD (Dynamic Stochastic Approximation) is a version of the SAMC
Algorithm that attempts to dynamically choose the modification factor
rather than relying on system dependent parameters such as $t_0$ or
$\gamma_0$.  There is an immediate advantage of such an algorithm where
parameters are chosen independent of system size or type. SAMC can be
applied as a dynamic importance sampling algorithm that uses an update
factor to explore energy space~\cite{liang2007stochastic,
werlich2015stochastic, schneider2017convergence}.  The update factor is
defined in the original implementation~\cite{liang2007stochastic} in
terms of an arbitrary tunable parameter $t_0$.
\begin{align}
\gamma_{t}^{\text{SA}} = \frac{t_0}{\max(t_0,t)}
\end{align}
We argue that it is possible to set this parameter to a system
independent parameter such that $t_0 = N_E^2/t$ where $N_E$ is the
number of energies that have been found.  One of the energy found terms
can be thought of as a change in the entropy $\Delta S$.  The change in
the entropy can be thought of as a slope of the triangle made by
connecting the minimum important energy and the maximum entropy state
divided by the minimum temperature:
\begin{align}
\Delta S = \frac{E_{\text{max}}-E_{\text{min}}}{T_{\text{min}}}
\end{align}
This gives an update factor as follows:
\begin{align}
\gamma_{t}^{\text{SA}} = \frac{N_E\Delta S}{t} =
\frac{N_E}{t}\frac{E_{\text{max}}-E_{\text{min}}}{T_{\text{min}}}
\end{align}
The update factor has the same $1/t$ behavior as the original SAMC
algorithm; however now every time a new energy is discovered during the
course of the simulation, the update factor experiences a jump.

\subsection{SA-TMMC Algorithm} We introduce a new algorithm consisting
of the merger of the Stochastic Approximation (SA) method proposed by
Liang et al.~\cite{liang2007stochastic} and the TMMC Algorithm with
suitable adaptations.  While the Stochastic Approximation method has
recently been implemented in comparison with the Wang-Landau
Algorithm~\cite{werlich2015stochastic, schneider2017convergence}, it
seems quite natural to create a new Monte-Carlo method which we will
call SA-TMMC.  Like in the original implementation, we define an update
factor where $t_0$ is the number of Monte-Carlo moves and $t$ is the
current iteration.
\begin{align} \gamma_{t}^{\text{SA}} =
\frac{t_0}{\max(t_0,t)}
\end{align}
One potential problem of merging SA and TMMC is that SA provides no
nice transition to the later method due to the lack of a
user-controlled cutoff.  Also for SA Monte-Carlo, the choice of
parameter $t_0$ proves difficult particularly for complex systems.
Finally, we enforce a minimum temperature probability which prevents
the system from sampling states below the minimum temperature. We
discuss here how we resolve all of the potential issues we encountered
when integrating both SA and TMMC.

\subsubsection{Define an adaptive parameter $t_0$}
It has been shown that if $F_{t}^{\text{SA}}$ satisfies two conditions
the estimator is proven to converge to the real
DOS~\cite{liang2006theory, liang2007stochastic}.
\begin{align}
\sum_{t=1}^\infty \gamma_{t}^{\text{SA}} = \infty \quad\textrm{and}\quad
\sum_{t=1}^\infty (\gamma_{t}^{\text{SA}})^\zeta < \infty
\end{align}
where $\zeta \in \{1,2\}$.  For practical implementation with TMMC, we
find it necessary to reset $t_0$ everytime a new energy state is found.
This ensures that all energy states are sampled before TMMC move
probabilities begin to be favored.

\subsubsection{Define a statistical weight to control the transition between SA and TMMC}
Our algorithm can alternate between SA and TMMC on the fly using
weights to determine the probability that a move between adjacent
energy states is allowed.  We define the transition probability
$\text{SA}_{\text{Weight}}$ as
\begin{align}
\delta_{\text{TMMC}}^2 =\frac{1}{n_\downarrow}+\frac{1}{n_\uparrow}, \quad\quad
\delta_{\text{SA}}^2 = \frac{1}{N_E}
\end{align}
\begin{align}
\text{SA}_{\text{Weight}} =\frac{\delta_{\text{TMMC}}^2}{\delta_{\text{TMMC}}^2 + \delta_{\text{SA}}^2}
\end{align}
where $N_\text{found}$ is the number of energy states that have been
found i.e. visited, and $n_{\uparrow\downarrow}$ are the number of
upward or downward moves between a pair of energy states respectively.
The rational behind the transition probability is that the transition
weight can be related to the fractional uncertainty in upward and
downward moves assuming there is no system correlation (the moves are
independent).  Although this is not necessarily the case, we find that
this definition for the transition weight is satifactory. The
probability of a move is then the geometic average of either using the
SA or TMMC Algorithm.

\begin{align}
\mathcal{P}_\text{move} = \mathcal{P}_\text{move}^{\text{SA}_\text{Weight}} \mathcal{P}_\text{move}^{1-\text{SA}_\text{Weight}}
\end{align}

\subsubsection{Enforce a minimum temperature probability}
One potentially damaging aspect to the convergence of the Stochastic
Approximation method is the need to enforce a minimum temperature
probability.  We define this probability as
\begin{align}
P_{\text{min}} = e^{\beta_{\text{max}}\Delta\epsilon}
\end{align}
where $\beta_{\text{max}} = \nicefrac{1}{k_B T_\text{min}}$.  If the
transition probability is ever below the minimum temperature transition
probability, we set the former equal to the later.

@@

For the square-well fluid, our simulations treat systems with a
well-width of $\text{ww} = 1.30$ and a filling fraction of either
$\text{ff} = 0.22$ or $\text{ww} = 0.30$.  The simulation explores the
energy space of the system down to a reduced temperature of
$T_{\text{min}} = 0.2$.  All simulations lead to the minimum important
energy $E_{\text{min}}$ and maximum entropy state $S_{\text{max}}$
being calculated (with the exception of the WL-TMMC method where both
of these parameters are needed a priori).  In the sections below, we
examine systems with different particles sizes, system dimensions, and
sticky-wall.

\subsubsection{Sticky-wall configuration}
A sticky-wall is an attractive wall that impacts particle moves.  Only
one such wall is inserted into a simulation and the intent is to assist
in visualizing phase separation.  As the temperature of the system
decreases, an increase in particle aggregation near the sticky-wall
becomes evident. As the energy of the particles is decreased, the
particles are unlikely to be able to escape the attractive sticky-wall
potential.  By examining the particle number density vs the system
length (z-axis), phase transitions to a liquid-vapor system from a
vapor dominated system become evident.

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/lv/ww1_30-ff0_22-30x4-max-entropy-error.pdf}
  \caption{The average entropy error for each method is shown with
  SAD serving as the `golden'
  standard.}\label{fig:avgS-30x4}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/lv/ww1_30-ff0_22-60x8-max-entropy-error.pdf}
  \caption{The average entropy error for each method is shown with SAD serving as
  the `golden' standard.}\label{fig:avgS-60x8}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/lv/ww1_30-ff0_22-100x10-max-entropy-error}
  \caption{The average entropy error for each method is shown with
    WL-TMMC using a cutoff of 10$^{-10}$ serving as the
    standard.}\label{fig:avgS-100x10}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_20-N50-max-entropy-error.pdf}
  \caption{The average entropy error is shown with TMI acting as the
  `golden' standard.}\label{fig:N50-mean-error}
\end{figure}
\begin{figure}
\includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N500-max-entropy-error.pdf}
  \caption{The average entropy error is shown with TMI acting as the
  `golden' standard.}
\end{figure}
